Active Learning from Blackbox to Timed Connectors

### 2016-01-14 paper/paper.tex
.

As a branch of \emph{machine learning} technique, active learning offers a way to obtain models from
low-level blackboxes. Works in \cite{DBLP:journals/mt/Daelemans10, DBLP:journals/iandc/Angluin87,
DBLP:conf/fase/RaffeltS06} shows impressive examples where active learning is used to extract
\emph{finite state automatas} or \emph{grammars}.

\todo{better expressions!}

@@

\begin{itemize}
  \item [-] A \emph{Sync} channel accepts a data item from its source end iff. the data item can be
    dispensed to its sink end simultaneously.
  \item [-] A \emph{LossySync} channel is always prepared to accept data items. These items will be
    send to its sink end simutaneously if possible, otherwise they will be dropped.
  \item [-] A \emph{SyncDrain} channel has two source ends and no sink end. It can accept a data
    item through one of its source end iff. a data item is also available for it to simultaneously
    accept through the other end. Then both two data items will be lost.
  \item [-] A A \emph{FIFO1} channel is an asychronous channel with one buffer cell. It accepts a
    data item whenever the buffer is empty.  
\end{itemize}

@@

\begin{definition}[Parameterized Mealy Machine]
  A \emph{Parameterized Mealy Machine} is defined as a 6-tuple $\mathcal{PM}=\langle
  S(\Sigma), s_0, I, O, \delta(\Sigma), \lambda(\Sigma)\rangle$ where 
  \begin{itemize}
    \item[-] $\Sigma$ is a \emph{finite} datum alphabet (hereinafter referred to as an alphabet)
    \item[-] $S$ is a function that maps an alphabet to a \emph{finite} set of
      states. We use $S(\Sigma)$ to denote the state set.
    \item[-] $I$ is a finite set of source-ends.
    \item[-] $O$ is a finite set of sink-ends.
    \item[-] $s_0$ is the initial state. It satisfies $\forall \Sigma,s_0\in S(\Sigma)$
    \item[-] $\delta$ maps a \emph{finite} datum alphabet to an \emph{output function}. We use
      $\delta(\Sigma):S(\Sigma)\times Input(\Sigma,I)\rightarrow Output(\Sigma, O)$ to denote the output function.
    \item[-] $\lambda$ maps an alphabet to a \emph{transition function}. We use
      $\lambda(\Sigma):S(\Sigma)\times Input(\Sigma,I)\rightarrow S(\Sigma)$ to denote the transiton
      function.
  \end{itemize}
\end{definition}

In the definition above, \emph{Input} and \emph{Output} are used to generate a set of input actions
and output actions from the corresponding alphabets and ends. For example, if we have a alphabet
consisting of two items $d_0$ and one source-end named $A$, the input actions would be
\begin{eqnarray*}
  Input(\{d_0\},\{A\}) &=& \{\{A:d_0,B:\rread\},\{A:d_0,B:\rnoread\}, \\
  & & \{A:\bot,B:\rread\},\{A:\bot,B:\rnoread\},T\}
\end{eqnarray*}
where we use $\rread$ to indicate that a end-sink is ready for write, and $\rnoread$ otherwise.

\todo{How to say?}
\[
Input(\Sigma,I)=\{\}\cup\{T\}
\]
where we use $T$ to denote the time action.
\[
Output(\Sigma,I)=\{\}
\]

@@

\begin{example}[PMM Semantics of FIFO1 channel]
  \label{example:pmmfifo}
  The semantics of FIFO channel with source end $A$ and sink end $B$ can be defined as follows.
  \begin{itemize}
    \item[-] $S(\Sigma)=\{q_0\}\cup\{q_d|d\in\Sigma\}$
    \item[-] $I=\{A\}$, $O=\{B\}$, $s_0=q_0$
    \item[-] output function
      \begin{displaymath}
        \delta(\Sigma)(s,i)=\left\{
        \begin{array}[h]{ll}
          (B:\bot) & s=q_0\land i=(A:\_,B:\rnoread) \\
          (B:d) & s=q_0\land i=(A:d,B:\rread)\\
          \rblock & s=q_0\land i=(A:\bot,B:\rread) \\
          (B:d) & s=q_d\land i=(A:\_,B:\rread)\\
          \rblock & s=q_d\land i=(A:d,B:\rnoread) \\
          (B:\bot) & s=q_d\land i=(A:\_,B:\rnoread) \\     
        \end{array}
        \right.
      \end{displaymath}
    \item[-] transition function
      \begin{displaymath}
        \lambda(\Sigma)(s,i)=\left\{
        \begin{array}[h]{ll}
          q_d & s=q_0\land i=(A:d,B:\rnoread) \\
          q_d & s=q_{d'}\land i=(A:d,B:\rread) \\
          q_0 & otherwise \\
        \end{array}
        \right.
      \end{displaymath}
  \end{itemize}
\end{example}

@@

\begin{example}[PMM Semantics of n-Timer channel]
  Considering a n-Timer channel with a source-end A and a sink-end B, we define its PMM-based
  semantics as:
  \begin{itemize}
    \item[-] $S(\Sigma)=\{q_0\}\cup\{q_d|d\in\Sigma\}$
    \item[-] $I=\{A\}$, $O=\{B\}$, $s_0=q_0$
    \item[-] output function
      \begin{displaymath}
        \delta(\Sigma)(s,i)=\left\{
        \begin{array}[h]{ll}
          (B:\bot) & s=q_0\land i=(A:\_,B:\rnoread) \\
          (B:d) & s=q_0\land i=(A:d,B:\rread)\\
          \rblock & s=q_0\land i=(A:\bot,B:\rread) \\
          (B:d) & s=q_d\land i=(A:\_,B:\rread)\\
          \rblock & s=q_d\land i=(A:d,B:\rnoread) \\
          (B:\bot) & s=q_d\land i=(A:\_,B:\rnoread) \\     
        \end{array}
        \right.
      \end{displaymath}
    \item[-] transition function
      \begin{displaymath}
        \lambda(\Sigma)(s,i)=\left\{
        \begin{array}[h]{ll}
          q_d & s=q_0\land i=(A:d,B:\rnoread) \\
          q_d & s=q_{d'}\land i=(A:d,B:\rread) \\
          q_0 & otherwise \\
        \end{array}
        \right.
      \end{displaymath}
  \end{itemize}
\end{example}

@@

\begin{definition}[Production of Parameterized Mealy Machines]
  Now we're going to define the production operator \emph{prod} of two parameterized mealy machines as,
  \[
  prod(PM_1,PM_2)=PM_3
  \]
  as follows. Here we assume that $PM_2.O\cap PM_1.I=\varnothing$
  \begin{itemize}
  	\item[-] $\forall\Sigma, PM_3.S(\Sigma)=PM_1.S(\Sigma)\times PM_2.S(\Sigma)$
    \item[-] $PM_3.I=PM_1.I\cup PM_2.I-PM_1.O$
    \item[-] $PM_3.O=PM_1.O\cup PM_2.O-PM_2.I$
    \item[-] $PM_3.s_0=(PM_1.s_0, PM_2.s_0)$
    \item[-] $\forall\Sigma, PM_3.\delta(\Sigma)((s_1,s_2), i)=$
      \begin{displaymath}
        \left\{
        \begin{array}[h]{ll}
          (Out_1 + Out_2)|_{PM_3.O} & \rblock\land Out_2\neq\rblock \\
          \rblock & otherwise \\
        \end{array}
        \right.
      \end{displaymath}
      where we have
      \begin{itemize}
        \item[*] $In_1 = i|_{PM_1.I}$
        \item[*] $Out_1 = PM_1.\delta(\Sigma)(s_1,In_1)$
        \item[*] $In_2 = (Out_1 + i)|_{PM_2.I}$
        \item[*] $Out_2 = PM_2.\delta(\Sigma)(s_2,In_2)$
      \end{itemize}
    \item[-] $\forall\Sigma, PM_3.\lambda(\Sigma)((s_1,s_2),i)=(s_1',s_2')$
      where we have
      \begin{itemize}
        \item[*] $s_1' = PM_1.\lambda(\Sigma)(s_1,In_1)$
        \item[*] $s_2' = PM_2.\lambda(\Sigma)(s_2,In_2)$
      \end{itemize}
  \end{itemize}
\end{definition}

\todo{example of production, maybe fifo+timer}

\section{From Blackbox to Timed Connectors} 
\label{sec:activelearning}
In this section, we will show how to use L* algorithm to extract models from blackboxes. Firstly, we
presents the definition of \emph{Observation Table}, which is used to describe the hypothesis.

@@

\subsection{Case Studies}
A simple example of timed connector is presented to show how L* works.
\begin{figure}[h]
  \begin{center}
    \input{./images/expfifo.tex}
  \end{center}
  \caption{Expiring FIFO1 Channel}
  \label{fig:expfifo}
\end{figure}

@@

\begin{table}[h]
  \renewcommand{\arraystretch}{1.3}
  \caption{Time-Cost Analysis}
  \label{tabel:timecost}
  \centering
  \begin{tabular}{l||rrr}
    \hline
    & FIFO & Alternator & Gate \\
    \hline\hline
    Membership Query(s) & 41.571 & 126.468 & 169.161 \\
    Hypothesis Query(s) & 0.001 & 0.003 & 0.004 \\
    Total Time(s) & 41.715 & 165.114 & 247.098\\
    Membership Query(\%) & 99.6 & 76.6 & 68.5\\
    \hline
  \end{tabular}
\end{table}

>>>>

As a branch of \emph{machine learning} technique, active learning offers a way to obtain models from
low-level models. Works in \cite{DBLP:journals/mt/Daelemans10, DBLP:journals/iandc/Angluin87,
DBLP:conf/fase/RaffeltS06} shows impressive examples where active learning is used to extract
\emph{Mealy machines} or \emph{regular languages} without time domain.

@@

\begin{itemize}
  \item [-] A \emph{Sync} channel accepts a data item from its source end iff. the data item can be
    dispensed to its sink end simultaneously.
  \item [-] A \emph{LossySync} channel is always prepared to accept data items. These items will be
    send to its sink end simutaneously if possible, otherwise they will be dropped.
  \item [-] A \emph{SyncDrain} channel has two source ends and no sink end. It can accept a data
    item through one of its source end iff. a data item is also available for it to simultaneously
    accept through the other end. Then both two data items will be lost.
  \item [-] A \emph{FIFO1} channel is an asychronous channel with one buffer cell. It accepts a
    data item whenever the buffer is empty.  
\end{itemize}

@@

\begin{definition}[Parameterized Mealy Machine]
  A \emph{Parameterized Mealy Machine} is defined as a 6-tuple $\mathcal{PM}=\langle
  S(\Sigma), s_0, I, O, \delta(\Sigma), \lambda(\Sigma)\rangle$ where 
  \begin{itemize}
    \item[-] $\Sigma$ is a \emph{finite} datum alphabet (hereinafter referred to as an alphabet)
    \item[-] $S$ is a function that maps an alphabet to a \emph{finite} set of
      states. We use $S(\Sigma)$ to denote the state set.
    \item[-] $I$ is a finite set of source-ends.
    \item[-] $O$ is a finite set of sink-ends.
    \item[-] $s_0$ is the initial state. It satisfies $\forall \Sigma,s_0\in S(\Sigma)$
    \item[-] $\delta$ maps a \emph{finite} datum alphabet to an \emph{output function}. We use
      $\delta(\Sigma):S(\Sigma)\times Input(\Sigma,I,O)\rightarrow Output(\Sigma,I,O)$ to denote the
      output function.
    \item[-] $\lambda$ maps an alphabet to a \emph{transition function}. We use
      $\lambda(\Sigma):S(\Sigma)\times Input(\Sigma,I,O)\rightarrow S(\Sigma)$ to denote the transiton
      function.
  \end{itemize}
\end{definition}

In the definition above, \emph{Input} and \emph{Output} are used to generate a set of input actions
and output actions from the corresponding alphabets and ends. The value of $Input(\Sigma,I,O)$ is
defined as a set of functions with on $I\cup O$ where $forall f\in Input(\Sigma,I,O)$, we have
\[
\forall i\in I, f(i)\in \Sigma\cup\{\bot\}\land \forall o\in
O,f(o)\in\{\rread, \rnoread\}
\]
where we use $\rread$ to indicate that a end-sink is ready for write, and $\rnoread$ otherwise. Note
that a $\bot$ means that there is \emph{no} data items on a source (or sink) end. In the same way,
$Output(\Sigma,I,O)$ is defined as a set of functions on domain $O$, with an additional symbol
$\rblock$ indicating an input failure. Similarly, $\forall f\in Output(\Sigma,I,O), f\neq \rblock$
implies
\[
\forall o\in O, f(o)\in \Sigma\cup\{\bot\}
\]

\begin{example}[Input and Output]
  If we have a simple alphabet with only one item $d_0$, a source-end named $A$, and a
  sink-end named $B$, the input actions would be
  \begin{small}
    \begin{eqnarray*}
      Input(\{d_0\},\{A\},\{B\}) &=& \{\{A:d_0,B:\rread\},\{A:d_0,B:\rnoread\}, \\
      & & \{A:\bot,B:\rread\},\{A:\bot,B:\rnoread\},T\}
    \end{eqnarray*}
  \end{small}
  and its output actions
  \[
  Output(\{d_0\},\{A\},\{B\}) =\{\{B:d_0\},\{B:\bot\}, \rblock\}
  \]
\end{example}

If there no data item written to all sink ends, we use $\varnothing$ to denote the empty output. For
example, in the case given above, $\{B:\bot\}$ can be briefly rewritten as $\varnothing$.

@@

\begin{example}[PMM Semantics of FIFO1 channel]
  \label{example:pmmfifo}
  The semantics of FIFO channel with source end $A$ and sink end $B$ can be defined as follows.
  \begin{itemize}
    \item[-] $S(\Sigma)=\{q_0\}\cup\{q_d|d\in\Sigma\}$
    \item[-] $I=\{A\}$, $O=\{B\}$, $s_0=q_0$
    \item[-] output function
      \begin{displaymath}
        \delta(\Sigma)(s,i)=\left\{
        \begin{array}[h]{ll}
          (B:\bot) & s=q_0\land i=(A:\_,B:\rnoread) \\
          (B:\bot) & s=q_d\land i=(A:\_,B:\rnoread) \\     
          (B:d) & s=q_0\land i=(A:d,B:\rread)\\
          (B:d) & s=q_d\land i=(A:\_,B:\rread)\\
          \rblock & s=q_d\land i=(A:d,B:\rnoread) \\
          \rblock & s=q_0\land i=(A:\bot,B:\rread) \\
        \end{array}
        \right.
      \end{displaymath}
    \item[-] transition function
      \begin{displaymath}
        \lambda(\Sigma)(s,i)=\left\{
        \begin{array}[h]{ll}
          q_d & s=q_0\land i=(A:d,B:\rnoread) \\
          q_d & s=q_{d'}\land i=(A:d,B:\rread) \\
          q_0 & otherwise \\
        \end{array}
        \right.
      \end{displaymath}
  \end{itemize}
\end{example}

@@

\begin{example}[PMM Semantics of n-Timer channel]
  Considering a n-Timer channel with a source-end A and a sink-end B, we define its PMM-based
  semantics as:
  \begin{itemize}
    \item[-] $S(\Sigma)=\{q_{i,d}|1\leq i\leq n, d\in \Sigma\}\cup\{q_0\}$
    \item[-] $I=\{A\}$, $O=\{B\}$, $s_0=q_0$
    \item[-] output function
      \begin{displaymath}
        \delta(\Sigma)(s,i)=\left\{
        \begin{array}[h]{ll}
          (B:d) & s=q_{n,d}\land i=(A:d,B:\rread)\\
          \rblock & s=q_{n,d}\:\land \\
          & (i=T\lor i=(A:\_,B:\rnoread) \\
          \rblock & s=q_{j,d}\land 1 \leq j < n\land \\
          & (i=(A:\_,B:\rread)\lor \\
          & i=(A:d,B:\_))\\
          (B:\bot) & otherwise\\
        \end{array}
        \right.
      \end{displaymath}
    \item[-] transition function
      \begin{displaymath}
        \lambda(\Sigma)(s,i)=\left\{
        \begin{array}[h]{ll}
          q_{j+1,d} & s=q_{j,d}\land i=T\land0 < j < n\\
          q_0 & s=q_{n,d}\land i=(A:\bot,B:\rread) \\
          q_{1,d} & s=q_{n,d'}\land i=(A:d,B:\rread) \\
          q_{1,d} & s=q_0 \land i=(A:d,B:\_) \\
          s & otherwise \\
        \end{array}
        \right.
      \end{displaymath}
  \end{itemize}
\end{example}

@@

\todo{check the timed behavior}
\begin{definition}[Production Operator]
  Now we're going to define the production operator \emph{prod} of two parameterized Mealy machines as,
  \[
  prod(PM_1,PM_2)=PM_3
  \]
  as follows. Here we assume that $PM_2.O\cap PM_1.I=\varnothing$
  \begin{itemize}
  	\item[-] $\forall\Sigma, PM_3.S(\Sigma)=PM_1.S(\Sigma)\times PM_2.S(\Sigma)$
    \item[-] $PM_3.I=PM_1.I\cup PM_2.I-PM_1.O$
    \item[-] $PM_3.O=PM_1.O\cup PM_2.O-PM_2.I$
  \end{itemize}
  \emph{(Here we assume that sink ends of $PM_1$ can be connected to source ends $PM_2$, but not
  vise versa)}
  \begin{itemize}
    \item[-] $PM_3.s_0=(PM_1.s_0, PM_2.s_0)$
    \item[-] $\forall\Sigma, PM_3.\delta(\Sigma)((s_1,s_2), i)=$
      \begin{displaymath}
        \left\{
        \begin{array}[h]{ll}
          (Out_1 + Out_2)|_{PM_3.O} & Out_1\neq\rblock\land Out_2\neq\rblock \\
          \rblock & otherwise \\
        \end{array}
        \right.
      \end{displaymath}
      where we have
      \begin{itemize}
        \item[*] $In_1 = i|_{PM_1.I}$
        \item[*] $Out_1 = PM_1.\delta(\Sigma)(s_1,In_1)$
        \item[*] $In_2 = (Out_1 + i)|_{PM_2.I}$
        \item[*] $Out_2 = PM_2.\delta(\Sigma)(s_2,In_2)$
      \end{itemize}
    \item[-] $\forall\Sigma, PM_3. \lambda(\Sigma)((s_1,s_2),i)=(s_1',s_2')$
      where we have
      \begin{itemize}
        \item[*] $s_1' = PM_1.\lambda(\Sigma)(s_1,In_1)$
        \item[*] $s_2' = PM_2.\lambda(\Sigma)(s_2,In_2)$
      \end{itemize}
  \end{itemize}
\end{definition}

The idea in \emph{production} is quite simple. The output of $PM_1$ would be provided as part of the
input of $PM_2$. Time actions T is only executed simutaneously betwenn $PM_1$ and $PM_2$.

As mentioned in the notation above, we cannot connect a sink end of $PM_2$ and a source end of
$PM_1$, which makes it difficult to construct connectors like alternator in Figure
\ref{fig:reoconnector}. Now we define a \emph{link} operation to connect sink ends and source ends
in the same connector.

\begin{definition}[Link Operator]
  The \emph{link} operation constructs connector by linking a sink end $OUT$ to a source end $IN$
  within the same connector.   
  \[
  PM' = link(PM, OUT, IN)
  \]
  Here we asuume that $IN\in PM.I$ and $OUT\in PM.O$.

  \begin{itemize}
  	\item[-] $\forall\Sigma, PM'.S(\Sigma)=PM.S(\Sigma)$
    \item[-] $PM'.I=PM.I-\{IN\}$
    \item[-] $PM'.O=PM.O-\{OUT\}$
    \item[-] $PM'.s=PM.s$
    \item[-] $\forall\Sigma, i\in Input(\Sigma,PM'.I,PM'.O)$, $PM'.\delta(\Sigma)(s, i)=$
      \begin{equation}
        \label{equ:existd}
        PM.\delta(\Sigma)(s,T)|_{PM'.O}
      \end{equation}
      if $i=T$. When $i$ is a non-time action, we need to check whether there's a data item
      $d\in\Sigma$
      satiefies
      \[
      PM.\delta(\Sigma)(s,i\cup\{IN:d\})|_{OUT} = d
      \]
      if \emph{true}, we have $PM'.\delta(\Sigma)(s, i)=$
      \[
      PM.\delta(\Sigma)(s,i\cup\{IN:d\})|_{PM'.O}
      \]
      otherwise $PM'.\delta(\Sigma)(s, i)=\rblock$
    \item[-] $\forall\Sigma, i\in Input(\Sigma,PM'.I,PM'.O).'$ the transition
      function can be defined as $PM.\lambda(\Sigma)(s,i)=$
      \begin{displaymath}
        \left\{
        \begin{array}[h]{ll}
          PM.\lambda(\Sigma,T) & i=T \\
          PM.\lambda(\Sigma,i\cup\{IN:d\}) & \exists d\in\Sigma \mbox{ satisfes Equation
          \ref{equ:existd}} \\
          s & otherwise \\
        \end{array}
        \right.
      \end{displaymath}
  \end{itemize}
\end{definition}

With \emph{link} and \emph{prop} defined, complex connectors can be formed by simple channels in
form of \emph{parameterized mealy machines}. It will be transformed into a concrete Mealy machine
once the alphabet $\Sigma$ is provided.

\section{From Blackbox to Timed Connectors} 
\label{sec:activelearning}
In this section, we will show how to use L* algorithm to extract models from blackboxes, which
mainly includes 3 steps: \emph{constructing hypothesis}, \emph{enclosing hypothesis} and
\emph{checking hypothesis}.

@@

\subsection{Case Studies}
A simple example of timed connector is presented to show how L* works.
\begin{figure}[h]
  \begin{center}
    \input{./images/expfifo.tex}
  \end{center}
  \caption{Expiring FIFO1 Channel (ExpFIFO1)}
  \label{fig:expfifo}
\end{figure}

Informally speaking, an expiring FIFO1 channel with \emph{timeout} $n$ is able to accept a data item
and stored it in the buffer cell for $n$ time units. If a read operation on $B$ is performed within
$n$ time units, it will obtain the data item successfully and clear the buffer. However, the data
item would be dropped if no read operation comes.

@@

\begin{table}[h]
  \renewcommand{\arraystretch}{1.3}
  \caption{Time-Cost Analysis}
  \label{tabel:timecost}
  \centering
  \begin{tabular}{l||rrrr}
    \hline
    & FIFO & Alternator & Gate & ExpFIFO1\\
    \hline\hline
    Membership Query(s) & 41.571 & 126.468 & 169.161 &\\
    Hypothesis Query(s) & 0.001 & 0.003 & 0.004 &\\
    Total Time(s) & 41.715 & 165.114 & 247.098 &\\
    Membership Query(\%) & 99.6 & 76.6 & 68.5 &\\
    \hline
  \end{tabular}
\end{table}
